Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-23 manuscript.Rmd
Background 2

Drawing on these findings, Lersch challenged the SDM and AUM as a “needless dichotomy” [-@lersch2023] and proposed a “Life Course Adaption Model” (LCAM) instead, which draws on the life course perspective and models personal culture over the course of a panel as a linear trajectory that varies for each respondent. Lersch’s work addresses two shortcomings of the AUM and SDM comparison. First, the AUM as outlined by Kiley and Vaisey assumes that change can only occur as a Markovian process, in which responses to personal culture items at time $t$ are a function only of responses at time $t-1$. But early life experiences might indirectly shape peoples’ personal culture as they move into new social roles or environments, even if they do not affect personal culture right away. For example, experiences in childhood might not affect views about family structure until much later in life when people create their own families. The LCAM allows responses at earlier points than $t-1$ to influence responses at time $t$.

Second, Kiley and Vaisey only adjudicated the AUM and SDM on three-wave panel data covering four years, which might not be a long enough time span to adjudicate the two models they outline. In contrast, Lersch tests the LCAM against the AUM and SDM on panels covering a much longer time frame (between 3 and 36 years) with more waves (between 3 and 18 waves), allowing for a greater possibility of detecting durable change.

In his tournament of models, Lersch finds broad support for the LCAM in data from five countries, with the LCAM favored over the AUM and SDM on 297 of 428 questions he analyzed and the SDM favored over the other two models on 112 items, with the remaining questions producing inconclusive results (no questions preferred the AUM). He concludes that “new experiences over the life course, which expose individuals to new social environments, including novel interactions, and different sets of resources, can persistently move individuals’ personal culture in novel directions” [@lersch2023: p. 243-244].

Despite differences in how these researchers interpret their findings, the empirical results are not far apart. Kiley and Vaisey [-@kiley2020] found that the majority of questions they explored favored the AUM, meaning they showed some evidence of durable change among adults, a finding broadly compatible with Lersch’s results. At the same time, while Lersch [-@lersch2023] found that the LCAM was preferred in the majority of cases he examined, durable change does not account for a large proportion of variance explained on these items. On average, people changed only about .07 standard deviations over 10 years. Again, this is consistent with Kiley and Vaisey’s assertion that durable change is often small. And a quarter of the items Lersch studied still preferred the SDM, suggesting that even with looser assumptions about the process of change, more cases, and more waves, many questions show no evidence of durable change. Relatedly, in most studies purporting to find evidence of durable change in peoples’ personal culture, the change is often quite small. For example, Brocic and Miles [-@brocic2021] find that completing graduate degrees in humanities, arts, and social sciences shift peoples’ moral relativism only about .2 standard deviations on average compared to people with no college degree, and this is the largest effect they identify. And studies of aggregate change show that, even on items where cohort effects dominate, there is always evidence that some people change over time [@vaisey2016]. In other words, despite different interpretations, the results of previous work are in many ways highly consonant.

>>>>

Drawing on these findings, Lersch challenged the SDM and AUM as a “needless dichotomy” [-@lersch2023]. Instead he proposed the “Life Course Adaption Model” (LCAM). This model draws on the life course perspective to model personal culture as a linear trajectory over the course of a panel that varies for each respondent. Lersch rectifies two shortcomings of the AUM and SDM. First, the AUM, as described by Kiley and Vaisey, posits that changes follow a Markovian process where personal culture responses at time $t$ rely solely on responses at time $t-1$. Howevre, earlier life experiences can (directly and indirectly) mold personal culture when transitioning to new social roles or into new environments, even if their initial impact is delayed. For example, childhood events might influence views on family structures later when individuals form their own families. The LCAM considers therefore considers influences from earlier than just $t-1$ on responses at time $t$.

Second, Kiley and Vaisey's analysis of the AUM and SDM is based on three-wave panel data over four years, which might not be extensive enough to properly adjudicate [COMPARE? WE ARE USING ADJUDICATING A LOT] the two models. Lersch, however, evaluates the LCAM against the AUM and SDM using panel data spanning a wider duration (from 3 to 36 years) and more waves (3 to 18), offering a better change to observe durable change.

In Lersch's tournament of models, the LCAM gains broad [EXTENSIVE?] support from data across five countries. Out of 428 questions, the LCAM was favored in 297 cases while the SDM was preferred in only 112. The rest did not yield inconclusive results, and notably, no questions favoured the AUM. He concludes that “new experiences over the life course [...] can persistently move individuals’ personal culture in novel directions” [@lersch2023: p. 243-244].

Despite differing interpretations of these scholars, their empirical results align closely. Kiley and Vaisey [-@kiley2020] discovered that most questions they investigated favored the AUM, indicating evidence of durable change among adults -- a finding largely in line with Lersch’s. However, while Lersch [-@lersch2023] found a preference for the LCAM in many cases, durable change only accounted for a minor proportion of variance explained. On average, individuals changed [CAN WE SAY SHIFTED TO INDICATE THE LINEAR DRIFT] about .07 standard deviations over 10 years, echoing Kiley's and Vaisey’s view that such changes are typically modest. Furthermore, a quarter of the items in Lersch's study favoured the SDM, suggesting that despite broader assumptions about change and more extensive data, many questions reveal no evidence of durable change. Likewise, in studies claiming evidence of durable change in personal culture, the magnitude is often minimal. For instance, Brocic and Miles [-@brocic2021] note that obtaining graduate degrees in humanities, arts, and social sciences only shift [NUDGES?] peoples’ moral relativism by about .2 standard deviations compared to non-degree holders [DELETED THE LAST SENTENCE PART HERE; FELT UNNECESSARY]. Moreover, studies of aggregate change demonstrate that even on questions where cohort effects prevail, there is always evidence that some individuals change over time [@vaisey2016]. In other words, despite different interpretations, the results of previous work are in many ways highly consonant.
